---
title: "HW3"
author: "Clara Bowman"
format: html
editor: visual
embed-resources: true
---

```{r}
library(dplyr)
library(knitr)
library(tidytext)
library(ggplot2)
library(textdata)
```

```{r}
pubmed <- read.csv('pubmed.csv')
```

# Text Mining

## 1. Most Common Words

### Number of Abstracts for Each Search Term

```{r}
table(pubmed$term)
```

There are 5 search terms with between 317 and 981 abstracts.

### Tokenize Abstracts

```{r}
tokens <- pubmed %>%
  unnest_tokens(token, abstract)
tokens %>% count(token, sort=TRUE) %>% head(10)
```

The most common tokens are stop words. 'covid' and '19' are the most common tokens that are not stop words.

### Top 5 words by search term

```{r}
tokens %>%
  group_by(term) %>%
  count(token) %>%
  top_n(5,n) 
```

Most of the top 5 words for each term are stop words. Covid includes 'covid' and '19' and 'prostate' and 'cancer' are among the top 5 words for prostate cancer.

## 2. Removing Stop Words

```{r}
tokens_no_stop <- tokens %>%
  anti_join(stop_words, by=c('token'='word')) 
tokens_no_stop %>%
  count(token, sort=TRUE) %>% head(10)
```

Stop words have been removed, so 'covid', '19', 'patients', 'cancer' and 'prostate' are the 5 most common words overall, and most words seem to be medically relevant.

### Top 5 words by search term (no stop words)

```{r}
tokens_no_stop %>%
  group_by(term) %>%
  count(token) %>%
  top_n(5,n) 
```

Almost all of the most common words are different now that stop words are removed. Each search term's top 5 most common words are specific to the disease. The only word that is common in multiple terms is 'patients'.

## 3. Tokenize into bigrams

```{r}
bigrams <- pubmed %>%
  unnest_ngrams(bigram, abstract, n=2) %>%
  count(bigram, sort=TRUE)
bigrams %>%
  top_n(n = 10, wt = n) %>%
  ggplot(aes(x = n, y = reorder(bigram, n), fill = bigram)) + 
  geom_col() +
  scale_fill_brewer(palette = "Paired") + 
  labs(title = "Top 10 Most Frequent Bigrams", y = "Bi-gram") + 
  theme(legend.position = 'none')
```

## 4. TF-IDF By Search Term

```{r}
tf_idf<-tokens %>%
  count(token,term) %>%
  bind_tf_idf(token, term, n) %>%
  arrange(term, desc(tf_idf))
  
tf_idf %>%
  group_by(term) %>%
  top_n(5, tf_idf)
```

This list does not include stop words. Each term's top 5 words seem specific to that term which makes sense as words that come up less frequently in other search terms are given a higher tf-idf.

# Sentiment Analysis

## 5. Using NRC Lexicon

```{r warning=FALSE}
nrc_counts <- pubmed %>%
  unnest_tokens(word, abstract) %>%
  inner_join(get_sentiments('nrc'), by='word') %>%
  count(term, sentiment, sort = TRUE)
nrc_counts %>% 
  group_by(term) %>%
  top_n(1, n)
```

"Negative" is the most common sentiment for meningitis and prostate cancer; "positive" is the most common sentiment for covid, cystic fibrosis and preclampsia.

```{r}
nrc_counts %>% 
  filter(sentiment!='positive' & sentiment!='negative') %>% 
  group_by(term) %>%
  top_n(1, n)
```

The most common sentiment excluding 'positive' and 'negative' is fear for prostate cancer, covid, and meningitis, anticipation for preeclampsia and disgust for cystic fibrosis.

## 6. Using AFINN Library

```{r}
afinn_abstract_avg <- pubmed %>%
  mutate(abstract_num = row_number()) %>%
  unnest_tokens(word, abstract) %>%
  inner_join(get_sentiments('afinn'), by='word') %>%
  group_by(abstract_num) %>%
  summarise(avg_sentiment=mean(value),
            term=first(term))
ggplot(afinn_abstract_avg, aes(x=term, y=avg_sentiment, fill=term)) +
  geom_boxplot() +
  labs(title='Average Abstract Sentiment by Search Term',x='Term',y='Average Sentiment') +
  scale_fill_brewer(palette = "Paired")+
  theme(legend.position = 'none')
```

All of the search terms except cystic fibrosis have an average sentiment across abstracts of slightly below zero. Cystic fibrosis stands out as the only mean above zero – abstracts for this term tend to be more positive than the rest. Additionally, prostate cancer's abstracts have a much tighter range of average sentiment values than other terms.
