---
title: "HW4"
author: "Clara Bowman"
format: html
editor: visual
---

```{r warning=FALSE}
library(RSQLite)
library(DBI)
library(ggplot2)
library(plotly)
library(microbenchmark)
library(parallel)
```

# HPC

## Question 1

```{r}
# running cumulative sum by row
fun2 <- function(mat) {
  n <- nrow(mat)
  k <- ncol(mat)
  ans <- mat
  for (i in 1:n) {
    for (j in 2:k) {
      ans[i,j] <- mat[i, j] + ans[i, j - 1]
    }
  }
  ans
}

fun2alt <- function(mat) {
  t(apply(mat, 1, cumsum))
}
```

#### Test function

```{r}
# Use the functions with this data
set.seed(2315)
dat <- matrix(rnorm(200 * 100), nrow = 200)
microbenchmark(
  fun2 = fun2(dat),
  fun2alt = fun2alt(dat),
  times = 1000
)
```

```{r}
result <- fun2(dat)
resultalt <- fun2alt(dat)
#should sum to zero if outputs are same
sum(result-resultalt) 
```

The two functions have the same output. The alternate (vectorized) function typically works 2-4 times faster than the original function. However, the max duration of the test is sometimes higher for the alternate function due to outlying high times (likely due to overhead and not computational limits).

## Question 2

```{r}
fun3 <- function(x) {
  out <- numeric(length(x))
  cumsum_x <- 0
  for (i in seq_along(x)) {
    if (x[i] == 0) {
      cumsum_x <- 0
    } else {
      cumsum_x <- cumsum_x + x[i] 
    }
    out[i] <- cumsum_x
  }
  out
}

fun3alt <- function(x) {
  cumsum <- cumsum(x)                
  reset <- cummax((x == 0) * cumsum) 
  cumsum - reset                      
}
```

```{r}
x <- sample(0:10, 20, replace = TRUE)
y <- sample(0:10, 100, replace = TRUE)
sum(fun3(x)-fun3alt(x))
sum(fun3(y)-fun3alt(y))
```

The outputs of the two functions are the same (the difference of the outputs are zero).

```{r}
microbenchmark(
  fun3 = fun3(x),
  fun3alt = fun3alt(x),
  times = 1000
)
microbenchmark(
  fun3 = fun3(y),
  fun3alt = fun3alt(y),
  times = 1000
)
```

The median fun3alt time is about 3-4 times less than the median fun3 time. fun3alt occasionly has very high outlying run times (once again likely due to overhead instead of computation) that can falsely inflate the mean time.

## Question 3

```{r}
sim_pi <- function(n = 1000, i = NULL) {
  p <- matrix(runif(n*2), ncol = 2)
  mean(rowSums(p^2) < 1) * 4
}

# Here is an example of the run
set.seed(156)
sim_pi(1000) # 3.132
set.seed(1231)
system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
  print(mean(ans))
})

```

```{r}
cl <- makeCluster(4)
clusterSetRNGStream(cl, 123)
clusterExport(cl, "sim_pi")
system.time({
  ans2 <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
  print(mean(ans))
})
stopCluster(cl)

```

The total elapsed time is about half when using 4 cores instead of 1.

#### Plot Estimate Distribution

```{r}
plot.new()
plot(density(ans), col = "tomato", lwd = 2, 
     main = "Density of Pi Estimates", xlab = "Estimate")

lines(density(ans2), col = "steelblue", lwd = 2, lty=2)

legend("topright",
       legend = c("Sequential", "Parallel"),
       col = c("tomato", "steelblue"), lwd = 2, lty=c(1,2))
```

Both functions follow a very similar distribution of estimates for pi. The estimates are normally distributed and centered around about 3.14 (true value of pi) and most estimates lie between 3.10 and 3.18.

#### Plot Time Distribution

```{r}
#Entire cell takes about 1 minute or so since each system tim is run 30 times
t1 <- replicate(30,system.time({
  ans <- unlist(lapply(1:4000, sim_pi, n = 10000))
})['elapsed'])
cl <- makeCluster(4)
clusterSetRNGStream(cl, 123)
clusterExport(cl, "sim_pi")
t2 <- replicate(30, system.time({
  ans2 <- unlist(parLapply(cl, 1:4000, sim_pi, n = 10000))
})['elapsed'])
stopCluster(cl)
```

```{r}
boxplot(t1, t2,
        names = c("Sequential", "Parallel"),
        col = c("tomato", "steelblue"),
        main = "Runtime Comparison",
        ylab = "Seconds")
```

In 30 pi estimation tests, parallel computing always takes less time than sequential. The sequential version takes on average about 0.8 to 0.9 secons whil the parallelized version takes about 0.4 seconds. Parallel computing also has less variation in elapsed time.

# SQL

## Setup

```{r}
# Initialize a temporary in memory database
con <- dbConnect(SQLite(), ":memory:")

# Read in tables
plots <- read.csv('plots.csv')
species <- read.csv('species.csv')
surveys <- read.csv('surveys.csv')

# Copy data.frames to database
dbWriteTable(con, "plots", plots)
dbWriteTable(con, "species", species)
dbWriteTable(con, "surveys", surveys)
```

## Question 4

```{sql, connection=con}
PRAGMA table_info(surveys)
```

```{sql, connection=con}
SELECT month, day, year, species_id, weight
FROM surveys
WHERE year = 1999
ORDER BY species_id
```

## Question 5

```{sql, connection=con}
SELECT species_id, AVG(weight) AS avg_weight, MIN(weight) as min_weight, MAX(weight) as min_weight
FROM surveys
GROUP BY species_id
```

```{sql, connection=con}
SELECT species_id, AVG(weight) AS avg_weight, MIN(weight) AS min_weight, MAX(weight) AS max_weight
FROM surveys
WHERE weight >= 5 AND weight <= 10
GROUP BY species_id

```

## Question 6

```{r}
result_q6 <- dbGetQuery(con, "SELECT surveys.species_id, year, species.species, COUNT(*) AS count, AVG(weight) AS avg_weight 
                              FROM surveys
                              INNER JOIN species ON surveys.species_id = species.species_id
                              GROUP BY surveys.species_id, year")
head(result_q6)
```

## Question 7

```{sql, connection=con}
PRAGMA table_info(species)
```

```{sql, connection=con}
SELECT taxa, COUNT(*) AS num_species
FROM species
GROUP BY taxa
HAVING num_species > 10
```

## Question 8

```{sql, connection=con}
PRAGMA table_info(plots)
```

```{r}
result_q8 <- dbGetQuery(con, "SELECT genus, surveys.plot_id, COUNT(*) AS count
FROM surveys
INNER JOIN species ON surveys.species_id = species.species_id
GROUP BY genus, surveys.plot_id
ORDER BY surveys.plot_id, count DESC")
head(result_q8)
```

## Question 9

```{sql, connection=con}
SELECT species.species, plots.plot_type, COUNT(*) AS COUNT
FROM surveys 
INNER JOIN species ON surveys.species_id = species.species_id
INNER JOIN plots ON surveys.plot_id = plots.plot_id
GROUP BY species.species, plots.plot_type
```

## Question 10

```{sql, connection=con}
SELECT species, AVG(weight) as avg_weight
FROM surveys
INNER JOIN species ON surveys.species_id = species.species_id
WHERE taxa = 'Rodent'
GROUP BY species
```

## Question 11

```{r}
p1 <- result_q8 %>%
  filter(plot_id <= 5) %>%
  ggplot(aes(x=factor(plot_id), y=count, fill=genus)) + 
  geom_bar(stat='identity', position = 'stack') +
  labs(x='Plot ID', y="Number of Samples", fill='Genus', title='Number of Samples per Genus in Plots 1-5')
ggplotly(p1)
```

## Question 12

```{r warning=FALSE}
result_q6 %>%
  plot_ly(x=~year, y=~avg_weight, color=~species_id, type='scatter',mode='lines', 
          connectgaps=TRUE, #Connect non sequential points 
          line = list(shape = "spline"), #Smooth lines to reduce noise
          text = ~paste("Species:", species, "<br>ID:", species_id),
          hovertemplate = paste("%{text}<br>","Year: %{x}<br>", "Avg Weight: %{y}<extra></extra>")) %>%
  layout(title = "Average Weight of Species over Time", xaxis = list(title = "Year"), yaxis = list(title = "Average Weight"))
```
