---
title: "Lab7"
format: html
editor: visual
---

```{r}
library(httr)
library(xml2)
library(stringr)
```

## **Question 1: How many Sars-Cov-2 papers?**

```{r}
# Downloading the website
website <- xml2::read_html("https://pubmed.ncbi.nlm.nih.gov/?term=sars-cov-2")

# Finding the counts
counts <- xml2::xml_find_first(website, "//span[@class='value']")

# Turning it into text
counts <- as.character(counts)

# Extracting the data using regex
totalcount <- stringr::str_extract(counts, "[0-9,]+")

# Removing any commas/dots so that we can convert to numeric
totalcount <- gsub(',', '', totalcount)

totalcount <- as.numeric(totalcount)
print(totalcount)
```

## **Question 2: Get article abstracts and authors**

```{r}
# read in text, each line is a separate character
abstracts <- readLines('abstract-sars-cov-2-set.txt', warn = FALSE)
# combine all text into one character
abstracts <- paste(abstracts, collapse='\n') 
# split the text whenever 3 new lines occur in a row (indicating two blank lines)
abstracts <- strsplit(abstracts,'\n\n\n')[1]
# replace any remaining "\n" symbols with spaces
abstracts <- gsub("\n", " ", abstracts)
# replace multiple spaces with single space
abstracts <- gsub("\\s+", " ", abstracts)
```

## **Question 3: Distribution of universities**

```{r}
library(stringr)
institution <- str_extract_all(
  abstracts,
  "\\b[A-Z][A-Za-z ]+(University|State University|University of [A-Z][A-Za-z ]+|Institute of [A-Z][A-Za-z ]+)"
  ) 
institution <- unlist(institution)
sort(table(institution), decreasing = TRUE)[1:10]
```

"The University" is included as a University name. These are likely Universities that follow the format: The University of... My regex looks for capital word(s) before or after certain phrases. This can be refined to include certain lowercase words. Additionally, specific medical schools belonging to a university wont necessarily be counted correctly.

## **Question 4: Make a tidy dataset**

```{r}
# read in text, each line is a separate character
abstracts <- readLines('abstract-sars-cov-2-set.txt', warn = FALSE)
# combine all text into one character
abstracts <- paste(abstracts, collapse = "\n")


# split the text whenever 3 new lines occur in a row (indicating two blank lines)
abstracts <- strsplit(abstracts, split = "\n\n\n")

abstracts = unlist(abstracts)

journal <- str_extract(abstracts, "(?<=\\d{1,3}\\.\\s).*?(?=\\.\\s\\d{4})")
journal = as.data.frame(journal)
head(journal)

titles <- sapply(abstracts, function(x){
  unlist(strsplit(x, split = "\n\n"))[2]
}, USE.NAMES = FALSE)

Titles = as.data.frame(titles)
```

```{r}
authors <- sapply(abstracts, function(x){
  unlist(strsplit(x, split = "\n\n"))[3]
}, USE.NAMES = FALSE)

authors = data.frame(authors)
head(authors)
```

```{r}
affiliations <- str_extract(abstracts, "(?s)\\nAuthor information:\\s*(.*?)\\n\\s*\\n")
```

```{r}
affiliations<- gsub("\n", " ", affiliations)

affiliations <- gsub(" +", " ", affiliations)

affiliations = as.data.frame(affiliations)

papers <- cbind(Titles, journal, authors, affiliations) 

papers <- papers %>%
  mutate(
    affiliation = str_sub(affiliations, start = 21, end = -1),
    affiliations = str_trunc(affiliation, width = 100),
    authors = str_trunc(authors, width = 100)) %>%
  select(titles, journal, authors, affiliations)
head(papers)
```
