---
title: "Lab8"
format: html
editor: visual
---

```{r}
library(microbenchmark)

```

## Problem 1: Vectorization

1.  This function generates an `n x k` dataset with all its entries drawn from a Poission distribution with mean `lambda`.

```{r fun_alt}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n){
    x <- rbind(x, rpois(k, lambda))    
  }
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n * k, lambda), nrow = n, ncol = k)
  
}
```

```{r}
x1 <- fun1()
x2 <- fun1alt()

dim(x1)
dim(x2)
```

Both are 100x4 matrices

```{r benchmarking}
microbenchmark(
  fun1 = fun1(1000,4,4),
  fun1alt = fun1alt(1000,4,4),
  times = 100
)
```

The alternate version takes about 1/20th the amount of time as the original.

2.  This function finds the maximum value of each column of a matrix.

    ```{r}
    # Data Generating Process (10 x 10,000 matrix)
    set.seed(1234)
    x <- matrix(rnorm(1e4), nrow=10)

    # Find each column's max value
    fun2 <- function(x) {
      apply(x, 2, max)
    }

    fun2alt <- function(x) {
      y = ncol(x)
      out = numeric(y)
      
      for (k in 1:y) {
        out[k] <- max(x[, k])
      }
      out
    }
    ```

```{r}
x11 = fun2(x)
x22 = fun2alt(x)

#Should sum to 0 if entries same
sum(x11-x22) 
```

```{r}
microbenchmark(fun2(x), fun2alt(x))
```

The alternative function is about 1/3 faster than the original

## Problem 2: Parallelization

1.  This function implements a serial version of the bootstrap. Edit this function to parallelize the `lapply` loop, using whichever method you prefer. Rather than specifying the number of cores to use, use the number given by the `ncpus` argument, so that we can test it with different numbers of cores later.

```{r}
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
  
  # THIS FUNCTION NEEDS TO BE PARALELLIZED
  # EDIT THIS CODE:
  ans <- parallel::mclapply(seq_len(R), function(i) {
    stat(dat[idx[, i], , drop = FALSE])
  }, mc.cores = ncpus)
  
  # Converting the list into a matrix
  ans <- do.call(rbind, ans)

  return(ans)
}
```

```{r}
# Bootstrap of an OLS
my_stat <- function(d) coef(lm(y ~ x, data=d))

# DATA SIM
set.seed(1)
n <- 500; R <- 1e4

x <- cbind(rnorm(n)); y <- x*5 + rnorm(n)

# Checking if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)

# You should get something like this
t(apply(ans1, 2, quantile, c(.025,.975)))
##                   2.5%      97.5%
## (Intercept) -0.1372435 0.05074397
## x            4.8680977 5.04539763
ans0
##                  2.5 %     97.5 %
## (Intercept) -0.1379033 0.04797344
## x            4.8650100 5.04883353
```

The result is very similar to the expected output (some variation is expected)

```{r}
system.time({my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 2L)})

system.time({my_boot(dat = data.frame(x, y), my_stat, R = R, ncpus = 1L)})

```

Using 2 cores made the total elapsed time about half (1.648 vs 3.020 seconds) which can have a very large impact when scaled to a larger dataset.

## Problem 3: Short Answer

1.  How you can decide if a problem is fit for parallelization?

    There are many tasks that can be completely *independently* from one another (or at least without each others outputs)

2.  How you can think about the relationship between vectorization and parallelization? How do we know whether we should use one or the other or both?

    Both vectorization and parallelization are methods to speed up processing. Vectorization applies the same operation to many data elements at once using optimized low-level code, while parallelization distributes independent tasks across multiple cores or machines. The choice depends on task structure: use vectorization when operations are uniform across data, parallelization when tasks can run independently, and combine both when possible.
